---
title: "Step1-Recurrence risk ratio to control group"
format: html
author: "Yiling Zhou & Yiran Li" 
date: 05-02-2024 
update date: 24-04-2024
R version: "4.2.3" 
editor: visual 
editor_options:
  chunk_output_type: console
---

## Load packages

```{r}
library(tidyverse) 
library(data.table) 
library(marginaleffects)
library(purrr)
library(speedglm)
library(foreach)
library(doParallel)
```

## Functions

### Function for estimating number of relatives affected vs 0

1.  marginal lambda
2.  marginal prevalence risk ratio (1 vs 0)

```{r}
# calculate RRR
getRRR <- function( Outcome, Exposure, Relative, dt)
  {
  RRR_var <- paste0(Exposure, "_", Relative)
  Outcome_var <- paste0(Outcome, "_individual")
  count_var <- paste0("counts_", Relative)
  birth_year_var <-  paste0("birth_year_", Relative)
  
  dt <- dt[!is.na(get(RRR_var)), ]
  dt[, age := givenYear - birth_year]
  # calculate relative age
  dt[, age_relative := givenYear - get(birth_year_var)]
  dt <- dt[age<=80 &age_relative<=80,]
  
  # categorize
  if(Relative == "Parent"){
    dt[, NumberAffected := get(RRR_var)]
    dt[, NumberAffected := as.character(NumberAffected)]
    dt[, NumberAffected:=factor(NumberAffected,
                                levels = c("0", "1", "2"))]
  }else{
    dt[, NumberAffected := get(RRR_var)]
    dt[NumberAffected>1, NumberAffected := 1]
    dt[, NumberAffected := as.character(NumberAffected)]
    dt[, NumberAffected:=factor(NumberAffected,
                                levels = c("0", "1"))]
  }
  
  dt[get(count_var)>2, (count_var) := 3]
  # categorize the kinship size
  if(max(dt[,get(count_var)]) >2){
    dt[, kinshipSize := cut(get(count_var),
                          breaks = c(1, 2, 3, Inf),
                          labels = c("1", "2", "Over2"),
                          right = FALSE)]
  }else{
      dt[, kinshipSize := as.character(get(count_var))]
  }
  
  # counts
  # interaction term in the model
  if(uniqueN(dt$kinshipSize) > 1){
    fitFormula <- as.formula(paste0(Outcome_var, " ~ NumberAffected*(age+ I(age^2) + sex)+age_relative+I(age_relative^2)+kinshipSize"))
  }else{
    fitFormula <- as.formula(paste0(Outcome_var, " ~ NumberAffected*(age+ I(age^2) + sex)+age_relative+I(age_relative^2)"))
  }
  
  # model
  model <- speedglm(fitFormula, data = dt,family = binomial())

  # avg_prediction calculates robust se for point in the prediction
  ## estimate prevalence in total population
  denominator <- avg_predictions(model, type = "response", vcov = TRUE)
  # counterfactual prediction for RRR_var = each level, same distribution as the empirical distribution of the total study population
  ## estimate marginal prevalence in each level of RRR_var
  numerator <- avg_predictions(model, variables = "NumberAffected", type = "response", vcov = TRUE) %>% 
    arrange(desc(NumberAffected))
 
  # estimate the marginal lambda (counterfactual outcome)
  lambda <- numerator$estimate/denominator$estimate
  lambda_SE <- getRRR_SE(x = numerator$estimate, x_SE = numerator$std.error,
                         y = denominator$estimate, y_SE = denominator$std.error)
  lambda_L <- lambda - qnorm(0.975)*lambda_SE
  lambda_H <- lambda + qnorm(0.975)*lambda_SE
  
  groupSize <- dt[, .N, by = "NumberAffected"] %>% 
    arrange(desc(NumberAffected))
  
  eventSize <- dt %>% 
    group_by(NumberAffected) %>% 
    count(!!sym(Outcome_var)) %>% filter(!!sym(Outcome_var)==1)%>% 
    arrange(desc(NumberAffected))

  # arrange output
  results <- tibble(
    Year = dt$givenYear[1],
    Outcome = Outcome,
    Exposure = Exposure,
    Relation = Relative,
    n_Relation = numerator[,1],
    Lambda = lambda,
    Lambda_SE = lambda_SE,
    Lambda_L = lambda_L,
    Lambda_H = lambda_H,
    PrevalenceR = numerator$estimate,
    PrevalenceR_SE = numerator$std.error,
    Prevalence = denominator$estimate,
    Prevalence_SE = denominator$std.error,
    EventSize = eventSize$n,
    GroupSize = groupSize$N,
    SampleSize = uniqueN(dt$RINPERSOON)
    ) 
  return(results)
}
```

## Get RRR with functions

### Labels

```{r}
exposures <- c("ADHD","ASD","T2D", "hypertension", "dyslipidemia", "stroke", "AP", "MI", "HF") 
outcomes <- c("ADHD", "ASD","T2D", "hypertension", "dyslipidemia", "stroke", "AP", "MI", "HF") 
relatives_fdr <- c("FDR") 
relatives_parent <- c("Parent", "Father", "Mother") 
relatives_sibling <- c("Sibling", "Brother", "Sister")  
relatives_child <- c("Offspring", "Son", "Daughter") 
relatives_auntUncle <- c("AuntUncle", "Uncle","Aunt")
relatives_NN <- c("NieceNephew","Nephew","Niece")
relatives_spouse <- c("Spouse", "Husband", "Wife")
relatives_grandparent <- c("Grandparent","Grandfather","Grandmother")
relatives_grandchild <- c("Grandchild")
relatives_halfsibling <- c("Halfsibling", "Halfbrother", "Halfsister")
relatives_sdr <- c("SDR") 
```

### Get Results

```{r}
parameters_dt_relative_newGeneration <- tibble(
   outcomes_vector = c(outcomes, outcomes[2:9], outcomes[3:9]),
   exposures_vector = c(exposures, rep("ADHD", 8), rep("ASD", 7))
 )
 
parameters_dt_relative_oldGeneration <- tibble(
  exposures_vector = c(exposures, exposures[2:9], exposures[3:9]),
  outcomes_vector = c(outcomes, rep("ADHD", 8), rep("ASD", 7))
)
#set cores for parallel
registerDoParallel(cores = 2)

# Use parent as an example
df <- fread("processData/phenotype_Parent_2020.csv")
parameters_dt <- parameters_dt_relative_oldGeneration

RRR_parent <- foreach(i = 1:length(relatives_parent), .combine = "rbind", 
                   .packages = c("purrr", "data.table", "speedglm",
                                 "marginaleffects", "tidyverse"),
                   .export = c("getRRR", "getRRR_SE"))%dopar%{
                  
                     relative <- relatives_parent[i]
                     
                     RRR <- map2_dfr(parameters_dt$outcomes_vector,
                                         parameters_dt$exposures_vector,
                                         ~getRRR(.x, .y, Relative = relative, 
                                                 dt = df))
                   }
saveRDS(RRR_parent, "processData/V2_RRR_results_parent_2020.rds")  

rm(df)
gc()
